Title Goes Here

### 2021-02-08 analysis/paper/paper.Rmd
Rotación y ENSO

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 43,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300
)

@@

@@

Figure \@ref(fig:cor-eof-naive-1) show the spatial pattern of the 4 leading EOF of zonal anomalies of geopotential height for 50 hPa and 200 hPa computed separately. The percentages XXX . At 50 hPa, EOF1 and EOF2 on one hand, and EOF3 and EOF4 on the other are clearly pairs of zonal waves with wave numbers 1 and 2-3 respectively shifted by 1/4 wavelength. That is, each group of two EOFs represents the same zonal wave structure that changes in magnitude and location. Similarly, at 200 hPa EOF2 and EOF3 represent the same zonal wave structure with wave number ~3.

@@

@@

ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30")
```
Both observations motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time. Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and argument of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the *navie* EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3. Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean. The real part of the first EOF is, therefore, capturing the mean zonally anomalous field *and the variability that projects onto it*.

@@

```{r extended-series, fig.cap = "(ref:extended-series-cap)"}

pattern <- ceof$eof[[1]]$right %>%
  sep_ReIm() %>%
  setnames("hgt", "EOF")

@@

@@

  geom_label(data = correlations[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)), 
             label.r = grid::unit(0, "lines"), label.size = 0) +
  geom_label(data = correlations[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)), 
             label.r = grid::unit(0, "lines"), label.size = 0) +

@@

  scale_x_datetime(NULL) +
  scale_y_continuous("Standarised EOF") +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C",
                                                           era5 = "ERA5 BE"),
                     guide = guide_legend(override.aes = list(size = 1), order = 1)) +
  facet_grid(part ~ PC, labeller = labeller(PC = lab_PC)) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +

  tag_facets("cr")
```

Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis) and using ERA20C for the period 1900 -- 2010. The hybrid ERA5 series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period. There is a slight downwards trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1" & dataset != "era20c"]$p_report` for the period 1950 -- 2010) that appears to be part of a long-term signal, judging by the statistically significant trend observed in the ERA20C reanalysis from 1900 to 2010 (p-value `r trends_all[part == "Real" & PC == "PC1" & dataset == "era20c"]$p_report`) with similar magnitude during the period of overlap.

@@

<!-- (ref:extended-magnitude-cap) Same as Figure \@ref(fig:extended-series) but for EOF magnitude defined as $\sqrt{\mathrm{Real}^2 + \mathrm{Imaginary}^2}$.  -->

<!-- ```{r extended-magnitude, fig.cap = "(ref:extended-magnitude-cap)"} -->
<!-- magnitudes <- extended %>%  -->
<!--   copy() %>%  -->
<!--   dcast(time + PC + dataset ~ part, value.var = "hgt") %>%  -->
<!--   .[, ref := time %between% common_range(time, dataset)] %>% -->
<!--   .[, hgt := sqrt(Real^2 + Imaginary^2)] %>%  -->
<!--   .[, hgt := (hgt - mean(hgt[ref]))/sd(hgt[ref]), by = .(PC, dataset)] -->

<!-- correlations_mag <- magnitudes %>%  -->
<!--   dcast(time + PC  ~ dataset, value.var = "hgt") %>%  -->
<!--   na.omit() %>% { -->
<!--     d <- . -->
<!--     rbind( -->
<!--   d[, .(correlation = cor(era20c, era5)),  -->
<!--     by = .(PC, time = ifelse(year(time) >= 1979,  -->
<!--                                    as.character(mean(periods[2:3])), -->
<!--                                    as.character(mean(periods[1:2]))))], -->

<!--     d[, .(correlation = cor(era20c, era5)),  -->
<!--     by = .(PC)] %>%  -->
<!--     .[, time := as.character(mean(periods[c(1, 3)]))] -->
<!--     ) -->
<!--   } %>%  -->

<!--   .[, time := lubridate::as_datetime(time)] -->

<!-- trends_mag <- magnitudes %>%  -->
<!--   .[time %between% common_range(time, dataset)] %>%  -->
<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
<!--   rm_intercept() %>%  -->
<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
<!--   .[, p_report := report_p(p_val)] -->

<!-- trends_all_mag <- magnitudes %>%  -->
<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
<!--   rm_intercept() %>%  -->
<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
<!--   .[, p_report := report_p(p_val)] -->

<!-- magnitudes %>%  -->
<!--   ggplot(aes(time, hgt)) + -->
<!--   # Zero line + rangeframe -->
<!--   geom_hline(yintercept = 0, size = 0.1, color = "gray30") + -->
<!--   ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") + -->

<!--   geom_line(aes(color = dataset)) + -->

<!--   # Correlations -->
<!--   annotate("errorbarh", xmin = periods[1:2], xmax = periods[2:3], y = -2.7, size = 0.3) + -->
<!--   annotate("errorbarh", xmin = periods[1], xmax = periods[3], y = -4, size = 0.3) + -->

<!--   geom_label(data = correlations_mag[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)),  -->
<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->
<!--   geom_label(data = correlations_mag[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)),  -->
<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->

<!--     geom_smooth(data = ~.x[dataset == "era20c"],  -->
<!--               method = "lm", se = FALSE,  aes(color = dataset)) + -->
<!--   geom_smooth(data = ~copy(.x)[time %between% common_range(time, dataset)], -->
<!--               aes(color = dataset), method = "lm", se = FALSE) + -->

<!--   scale_x_datetime(NULL) + -->
<!--   scale_y_continuous("Standarised EOF magnitude") + -->
<!--   scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C", -->
<!--                                                            era5 = "ERA5 BE"), -->
<!--                      guide = guide_legend(override.aes = list(size = 1), order = 1)) + -->
<!--   facet_wrap(~PC, labeller = labeller(PC = lab_PC), ncol = 2) + -->
<!--   theme(panel.spacing = grid::unit(1.5, "lines"))  + -->

<!--   tag_facets() -->
<!-- ``` -->

@@

```{r compute-regression}
regr <- ceof[, eof[[1]]$left, by = season] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = as.numeric(scale(R)), I = as.numeric(scale(I))), by = .(season, PC)] %>%
  dcast(time ~ PC, value.var = c("R", "I")) %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(850, 700, 200, 50)] %>%
  # .[variable == "hgt"] %>%
  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE), by = .(lev, lon, lat, variable)] %>%
  .[, c("term", "PC") := tstrsplit(term, "_")] %>%
  rm_intercept() %>% 
  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
```

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt") 
```
Figure \@ref(fig:ref:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.

@@

* Parte real de EOF1: variabilidady que se proyecta en la dirección del campo medio

* Parte imaginaria de EOF1: variabilidad asociada a la parte simétrica del SAM

* EOF2: PSA1 y PSA2.  
- Parte imaginaria: proyecta sobre la parte asimétrica del SAM (PSA2)
- Parte real: relativamente independiente del SAM

@@

@@

  ceof[, eof[[1]]$left] %>%
    sep_ReIm() %>%
    .[, hgt := scale(hgt), by = .(PC, part)] %>%
    index[., on = "time"] %>%
    na.omit() %>%
    .[, index := get(index_col)] %>%
    ggplot(aes(hgt, index)) +
    geom_point(aes(color = part)) +
    geom_smooth(aes(color = part), method = "lm") +
    shadowtext::geom_shadowtext(data = ~.x[, .(cor = cor(index, hgt),
                                               x = -1.2,
                                               y = sign(-cor(index, hgt))*.1 + predict(lm(index ~ hgt),
                                                                                       newdata = data.frame(hgt = -1.2))),
                                           by = .(PC, part)],
                                bg.colour = "white", bg.r = 0.2,
                                size = 5,
                                aes(x, y, color = part, label = paste0("r² = ", round(cor^2, 2)))) +
    scale_x_continuous("EOF") +
    scale_y_continuous(toupper(index_col)) +
    scale_color_brewer("Part", palette = "Dark2") +
    facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2")))
}
```

```{r sam-eof, fig.cap = "Standarised temporal EOFs and SAM index computed as the leading EOF of geopotential height anomalies at 700 hPa."}
plot_with_index(ceof, sam[lev == 700 & symmetry == "full"][, .(time, sam)])
```

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM and Symmetric SAM indices computed at each level according to Campitelli on press??.

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
time <- ceof[, eof[[1]]$left] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)] %>%
  .[]

sams <- readRDS(data_path("derived", "sam.Rds"))

sams %>% 
  .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>%
  .[, cor(hgt, estimate)^2, by = .(lev, term, part, PC)] %>%
  ggplot(aes(lev, V1)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²") +
  scale_size_manual(guide = "none",
                    values = c(full = 1,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "Asymmetric SAM",
                                                         sym = "Symmetric SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip()

@@

>>>>

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 43,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

Whereas for traditional "real" PCA, each principal component is defined up to it's sign. In complex PCA, principal components are defined up to their phase [@horel1984].

@@

Figure \@ref(fig:cor-eof-naive-1) show the spatial pattern of the 4 leading EOF of zonal anomalies of geopotential height for 50 hPa and 200 hPa computed separately. The percentages XXX . At 50 hPa, EOF1 and EOF2 on one hand, and EOF3 and EOF4 on the other are clearly pairs of zonal waves with wave numbers 1 and 2-3 respectively shifted by 1/4 wavelength. That is, each group of two EOFs represents the same zonal wave structure that changes in magnitude and location. Similarly, at 200 hPa EOF2 and EOF3 represent the same zonal wave structure with wave number \~3.

@@

```{r rotate_eofs}
rotate <- function(z, angle = 0) {
  complex(real = cos(angle), imaginary = sin(angle)) * z
}

box <- list(lon = c(150, 195),
           lat = c(-25, -5))

psi_box <- psi %>% 
  .[lev == 200] %>% 
  .[lat %between% box$lat & lon %between% box$lon] %>% 
  .[, .(psi.z = weighted.mean(psi.z, cos(lat*pi/180))), by = .(time, lev)]

psi_box <- cut(ceof$eof[[1]], 2)$left %>% 
  copy() %>% 
  .[psi_box, on = "time"]

angles <- seq(-pi, pi, by = 1*pi/180)

rotations_pc1 <- lapply(angles, function(a) {
  ceof$eof[[1]]$left[PC == "PC1"] %>% 
    copy() %>% 
    .[, hgt := rotate(hgt, a)] %>% 
    sep_ReIm() %>% 
    .[, .(mean = mean(hgt)), by = part] %>%
    .[, rotation := a]
}) %>% 
  rbindlist()

best_rotation_pc1 <- rotations_pc1[part == "Imaginary"][which.min(abs(mean))]$rotation

rotations_pc2 <- lapply(angles, function(a) {
  psi_box %>% 
    .[, hgt2 := rotate(hgt, a)] %>% 
    .[, .(R = cor(Re(hgt2), psi.z),
          I = cor(Im(hgt2), psi.z))] %>% 
    .[, rotation := a]
}) %>% 
  rbindlist()

best_rotation_pc2 <- rotations_pc2[which.min(abs(R^2))]$rotation

eof_rotated_left <- ceof$eof[[1]]$left %>% 
  copy() %>% 
  .[PC == "PC2", hgt := rotate(hgt, best_rotation_pc2)] %>% 
  .[PC == "PC1", hgt := rotate(hgt, best_rotation_pc1)]

eof_rotated_right <-  ceof$eof[[1]]$right %>% 
  copy() %>% 
  .[PC == "PC2", hgt := rotate(hgt, best_rotation_pc2)] %>% 
  .[PC == "PC1", hgt := rotate(hgt, best_rotation_pc1)]

ceof_unrotated <- copy(ceof)

ceof$eof[[1]]$left <- eof_rotated_left
ceof$eof[[1]]$right <- eof_rotated_right

ceof <- rbindlist(list(rotated = ceof,
               unrotated = ceof_unrotated), 
          idcol = "rotation")

rotated <- function(x) x[rotation == "rotated"] 
unrotated <- function(x) x[rotation == "unrotated"] 
```

```{r check-rotation}
# Check that rotated and unrotated eofs 
# predict the same fields (up to floating point errors)
bad_cells <- ceof[, predict(eof[[1]]), by = rotation] %>% 
  .[, hgt := as.numeric(Re(hgt))] %>% 
  dcast(time + lon + lat + lev ~ rotation, value.var = "hgt") %>% 
  .[, dif := rotated - unrotated] %>% 
  .[abs(rotated - unrotated)  > 1e-10]

if (nrow(bad_cells) != 0) {
  stop("Rotated and unrotated eofs are not equivalent!")
}

ceof <- ceof[rotation == "rotated"][, rotation := NULL][]
```

@@

ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30")
```

Both observations motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time. Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and argument of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the navie EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3. Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean. The real part of the first EOF is, therefore, capturing the mean zonally anomalous field and the variability that projects onto it.

@@

```{r extended-series, fig.cap = "(ref:extended-series-cap)"}
pattern <- ceof$eof[[1]]$right %>%
  sep_ReIm() %>%
  setnames("hgt", "EOF")

@@

# Check that the extended index coincides with the 
# original index.
cors <- era_more %>% 
  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>% 
  na.omit() %>% 
  .[, cor(hgt, i.hgt), by = .(PC, part)]

if (any(cors$V1 < 0.98)) {
  stop("Extended indices didn't work!")
}

@@

  geom_label(data = correlations[time %in% range(time)], 
             aes(y = -2.7, label = paste0("r²: ", signif(correlation^2, 2))), 
             label.r = grid::unit(0, "lines"), label.size = 0) +
  geom_label(data = correlations[!(time %in% range(time))], 
             aes(y = -4, label = paste0("r²: ", signif(correlation^2, 2))), 
             label.r = grid::unit(0, "lines"), label.size = 0) +

@@

  scale_x_datetime(NULL, date_labels = "%Y",
                   breaks = function(x) sort(unique(c(scales::pretty_breaks()(x), 
                                                          lubridate::as_datetime("1979-10-01"))))) +
  scale_y_continuous("Standarised EOF") +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C",
                                                         era5 = "ERA5 BE"),
                     guide = guide_legend(override.aes = list(size = 1), order = 1)) +
  facet_grid(part ~ PC, labeller = labeller(PC = lab_PC)) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  tag_facets("cr")
```

Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis) and using ERA20C for the period 1900 -- 2010. The hybrid ERA5 series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period. There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1" & dataset != "era20c"]$p_report` for the period 1950 -- 2010) that appears to be part of a long-term signal, judging by the statistically significant trend observed in the ERA20C reanalysis from 1900 to 2010 (p-value `r trends_all[part == "Real" & PC == "PC1" & dataset == "era20c"]$p_report`) with similar magnitude during the period of overlap.

@@

@@

```{r compute-regression}
regr <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = as.numeric(scale(R)), I = as.numeric(scale(I))), 
    by = .(season, PC)] %>%
  dcast(time ~ PC, value.var = c("R", "I")) %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(850, 700, 200, 50)] %>%
  # .[variable == "hgt"] %>%
  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE), 
    by = .(lev, lon, lat, variable)] %>%
  .[, c("term", "PC") := tstrsplit(term, "_")] %>%
  rm_intercept() %>% 
  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
```

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt") 
```

Figure \@ref(fig:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.

@@

-   Parte real de EOF1: variabilidady que se proyecta en la dirección del campo medio

-   Parte imaginaria de EOF1: variabilidad asociada a la parte simétrica del SAM

-   EOF2: PSA1 y PSA2.\

-   Parte imaginaria: proyecta sobre la parte asimétrica del SAM (PSA2)

-   Parte real: relativamente independiente del SAM

@@

### SAM

@@

  ceof[, eof[[1]]$left] %>%
    sep_ReIm() %>%
    .[, hgt := scale(hgt), by = .(PC, part)] %>%
    index[., on = "time"] %>%
    na.omit() %>%
    .[, index := get(index_col)] %>%
    ggplot(aes(hgt, index)) +
    geom_point(aes(color = part)) +
    geom_smooth(aes(color = part), method = "lm") +
    shadowtext::geom_shadowtext(data = ~.x[, .(cor = cor(index, hgt),
                                               x = -1.2,
                                               y = sign(-cor(index, hgt))*.1 + predict(lm(index ~ hgt),
                                                                                       newdata = data.frame(hgt = -1.2))),
                                           by = .(PC, part)],
                                bg.colour = "white", bg.r = 0.2,
                                size = 5, show.legend = FALSE,
                                aes(x, y, color = part, label = paste0("r² = ", round(cor^2, 2)))) +
    scale_x_continuous("EOF") +
    scale_y_continuous(toupper(index_col)) +
    scale_color_brewer("Part", palette = "Dark2") +
    facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2")))
}
```

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2020a. Points mark estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
time <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)] %>%
  .[]

sams <- readRDS(data_path("derived", "sam.Rds")) %>% 
  .[season(time) == "SON"] %>%
  .[, dim :=  lubridate::days_in_month(time)] %>% 
  .[, .(estimate = weighted.mean(estimate, dim)),
    by = .(lev, term, time = seasonally(time))]

sam_r2 <- sams %>% 
  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, with(cor.test(hgt, estimate), 
           list(estimate = estimate, 
                low = conf.int[1],
                high = conf.int[2],
                p_value = p.value)), by = .(lev, term, part, PC)] %>%

  .[, low2 := pmin(low^2, high^2, ifelse(sign(high) != sign(low), 0, 1))] %>% 
  # .[low2 <= 0, low2 := 0] %>% 
  .[, high2 := pmax(low^2, high^2)] %>% 
  .[PC != "PC3"]

max_r2 <- sam_r2[part == "Imaginary" & term == "asym"][, .SD[which.max(estimate^2)],
                                                       by = .(PC)]

sam_r2 %>% 
  ggplot(aes(lev, estimate^2)) +
  # geom_ribbon(data =  ~.x[p.adjust(p_value, "fdr") < 0.001], 
  # aes(ymin = low2, ymax = high2, color = term), alpha= 0.1) +
  geom_point(data = ~.x[p.adjust(p_value, "fdr") < 0.01], aes(color = term)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²", limits = c(0, 1)) +
  scale_size_manual(guide = "none",
                    values = c(full = 1.2,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "A-SAM",
                                                         sym = "S-SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip() +
  tag_facets("rc")
```

Figure \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices. The SAM index represents the leading EOF of monthly geopotential height fields south of 20ºS at each level. The A-SAM and S-SAM indices represent, respectively, the zonally asymmetric and symmetric component of the SAM and are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields (@campitelli2020a). Since the analysis here is only for the SON trimester, monthly values were averaged across semesters (weighted by the number of days in each month).

Both EOFs bear some modest relationship with the SAM index (thick green line in Figure \@ref(fig:sam-eof-vertical)), although not statistically significant at every level. The split between A-SAM and S-SAM gives more insight into the nature of the relationship. The relationship between the SAM and the imaginary EOF1 (Figure \@ref(fig:sam-eof-vertical).b1) is mediated by S-SAM in the troposphere, but by the A-SAM in the stratosphere.

The Imaginary EOF2 is related with the SAM through the A-SAM in the troposphere, with up to `r scales::percent(max_r2[PC == "PC2"]$estimate, 2)` of shared variance, reached at `r max_r2[PC == "PC2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2). The Imaginary EOF2 is, then, an almost a pure representation of the tropospheric asymmetric part of the SAM. Since the Real EOF2 represents a quarter-wavelength rotation of the same zonal wave 3 structure, the Imaginary part is the phase that coincides with the SAM. As a corollary, this means that the asymmetric component of the SAM represents only one phase of a zonally varying zonal wave.

```{r regr_psi}
regr_psi <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(psi.z, R, I, se = TRUE),by = .(lon, lat, lev, PC, season)] %>%
  rm_intercept() %>%
  .[, psi.z := estimate] %>%
  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)] %>% 
  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"),
                          ordered = TRUE)]
```

```{r regr_sst}
regr_sst <- ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(t, I, R, se = TRUE), by = .(lon, lat, PC, season)] %>%
  rm_intercept() %>% 
  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"),
                          ordered = TRUE)]
```

```{r plot_sst_psi}
plot_sst_psi <- function(which_pc, which_lev) {
  sst_gdata <- regr_sst[PC == which_pc] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    .[, var := "SST"]

  regr_psi %>% 
    .[PC == which_pc] %>%
    .[lev == which_lev] %>%
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    .[, var := "Streamfunction"] %>% 
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),
                      breaks = AnchorBreaks(exclude = 0)) +
    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 1)),
                geom = "point", size = 0.2, alpha = 0.5) +
    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
                    size = 0.2, res = 2) +
    geom_qmap(keep = .015) +
    scale_fill_divergent_discretised("Streamfunction", guide = "none") +
    ggnewscale::new_scale_fill() +

    geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_tanaka(data = sst_gdata, 
                        aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour2(data = sst_gdata, 
                  aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(data = sst_gdata,
                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 3)),
                geom = "point", size = 0.2, alpha = 0.5) +
    scale_fill_divergent_discretised("SST", guide = guide_colorsteps_bottom(order = 2)) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ var, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
    coord_quickmap(ylim = c(NA, 10)) +
    theme(legend.box = "horizontal") +
    # ggh4x::facet_nested(term ~ PC) +
    axis_labs_smol +
    tag_facets("cr")
}
```

```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
plot_sst_psi(which_pc = "PC2", which_lev = 200) 
# +   annotate("rect", xmin = 170, xmax = 195, ymin = -25, ymax = -5, color = "black", fill = NA)
```

??? -\> esta intro hay que cambiarla. To help identify those sources Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and STT and Streamfunction at 200 hPa. As expected from the correlations between each component of EOF2 and ENSO, Imaginary EOF2 is associated with strong ENSO-like SST anomalies and streamfunction (Figure \@ref(fig:sst-psi-2).a2). Streamfunction anomalies (Figure \@ref(fig:sst-psi-2).b2) are indicative of tropical forcing, which is compatible with SST anomalies. Real EOF2, on the other hand, is not associated with significant SST anomalies (Figure \@ref(fig:sst-psi-2).a1), as expected by its lack of relationship with ENSO, and streamfunction anomalies in the tropics are much weaker and non significant.

```{r enso_cor}
enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
    sep_ReIm() %>%
    .[, hgt := scale(hgt), by = .(PC, part)] %>%
    enso[., on = "time"] %>%
    na.omit() %>% 
   .[, with(cor.test(hgt, oni), list(estimate = estimate,
                                     p.value = p.value,
                                     low = conf.int[1],
                                     hig = conf.int[2])), 
     by = .(PC, part)] %>% 
  .[, cor_text := paste0(signif(estimate, 2), " (CI: ", 
                         signif(low, 2), " -- ", signif(hig, 2), ")")]

@@

# Appendix

## Chosen rotations of the EOFs

```{r rotations-pc1, fig.cap = "Rotations of PC1"}
rotations_pc1 %>% 
  ggplot(aes(rotation, mean)) +
  geom_hline(yintercept = 0, color = "gray") +
  geom_line(aes(color = part), size = 1) +
  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 30*pi/180), 
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous("Mean value") +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx)
```
